Maximizing Ink in Symmetric Partial Edge Drawings of $k$-plane Graphs

### 2018-01-08 eurocg/paper.tex
sec 4.2 alpha

In this section we close the gap between the known hardness of \maxsped~\cite{b-sgh-15} and the polynomial-time algorithm for 2-plane drawings~\cite{bcgkmn-pped-17} as summarized in the following theorem.

@@

\begin{proof}
	We reduce from the \NP-hard problem \ppsat~\cite{l-pftu-82} using similar ideas as Bruckdorfer's~\cite{b-sgh-15} general proof of the hardness of \maxsped. %, which reduces from the more restricted problem \textsc{Planar Positive 1-in-3 Sat}. 
	Let $\phi$ be a planar 3-Sat formula with $n$ variables $\{x_1, \dots, x_n\}$ and $m$ clauses $\{c_1, \dots, c_m\}$, each consisting of three literals.
	We can assume that $\phi$ comes with a planar drawing of its variable-clause graph $H_\phi$ that has a vertex for each variable $x_i$ placed on a horizontal line and a vertex for each clause $c_j$ connected on one side of the horizontal line to the three variables appearing in $c_j$. 
	In our gadget-based reduction we create a 3-plane drawing $\Gamma_\phi$ as a set of line segments of uniform length and a value $L$ such that $\Gamma$ has a SPED of total ink at least $L$ if and only if $\phi$ is satisfiable.

	All segments are of length 5.
	We construct our gadgets using pairs of intersecting segments, one colored green, the other one red.
	Their mutual intersection point is at distance 1 from an endpoint, which has the effect that the maximum amount of ink contributed by such a pair is 7 (one full segment and the other one with two stubs of length 1 each).

@@

\end{proof}

\section{Two polynomial special cases}

@@

Section~\ref{sec:hardness} showed that \maxsped is generally \NP-hard for $k \ge 3$. In this section we consider the special case that the intersection graph of the $k$-plane input drawing is a tree or a cactus. Since every planar graph is representable as intersection graph of line segments~\cite{chalopin2009every}. we can also encounter every tree or cactus. In both cases we present polynomial-time dynamic programming algorithms.

Let $ C = (V,E) $ be the intersection graph of the edge set $S$ in a given drawing $\Gamma$ of graph $ G $ as defined in Section~\ref{sec:preliminaries}. Let $ u \in V $ and $ \deg(u) = k $, then for the corresponding edge $ s(u) \in S $ there are $ k + 1 $ possible stub pairs including drawing the whole edge. We store the lengths of the possible stubs as node weights $ \ell_1(u), \dots, \ell_k(u) \in \mathbb{R}_+$ and assume that $ \ell_1(u),\dots,\ell_k(u) $ are sorted from shorter to longer stubs. We define  $ \ell_0(u) $ as the length of the whole edge $ s(u) $.

@@

\subsection{Trees}
\label{sec:tree}
Here we assume that the intersection graph $ C = (V,E) $ is a rooted tree of maximum degree $ k $. We give a bottom up dynamic programming algorithm for solving \maxsped on $ C $. For a vertex $ u \in V $ let $ p(u) $ denote the parent of $ u $ in $ C $ and let $ c(u) $ denote the set of its children. The ink values of the partial \maxsped solutions for the subtrees $T_u$ rooted at each vertex $ u \in V $ such that stub length $ \ell_i(u) $ is used for $s(u)$ are computed and stored as $ T_i(u) $.

\begin{figure}[tbp]
	\centering
	\includegraphics{two_solutions_needed}
	\caption{Consider $ u $, then only the blue and the green solution are relevant.}
	\label{fig:two_solutions}
\end{figure}

Since we know $ C $ is a tree every vertex $ u \in V $ with degree $ \delta $ has a parent $ p(u) \in V $. For this parent there exists one stub length $ \ell_j(p(u)) $ in the sequence of stub lengths $ \ell_0(p(u)),\dots,\ell_{\delta}(p(u)) $ such that drawing $ p(u) $ with these stubs does not intersect a completely drawn $ u $, while drawing $ p(u) $ with $ \ell_{j + 1}(p(u)) $, i.e. the next longer stub length, intersects $ \ell_0(u) $. We call the solution which would intersect the parent $ \sollong(u) $ and the one not intersecting the parent $ \solshort(u) $. See Figure~\ref{fig:two_solutions} for an illustration. More formally let $ j $ be the index such that stubs of length $ \ell_j(p(u)) $ do not intersect $ s(u) $, but $ \ell_{j +1}(p(u)) $ would, we set:
\begin{align*}
	\solshort(u) = \max\{T_1(u),\dots,T_j(u)\} \\
	\sollong(u) = \max\{T_0(u),\dots,T_{\delta}(u)\}
\end{align*}
It remains to define $ T_i(u) $ for a vertex $ u $ and stub length $ \ell_i(u) $:
\begin{align}
	\label{rec:tree}
	T_i(u) = \ell_i(u) + \sum_{v \in c(u)}
	\begin{cases}
		\solshort(v) & \text{if } s(u) \text{ with length } \ell_i \text{ intersects } s(v) \\
		\sollong(v) & \text{else}
	\end{cases}
\end{align}
Solving Recurrence~\ref{rec:tree} yields an optimal solution to the \maxsped problem on $ G $ with drawing $ \Gamma $ as the following lemma shows.

\begin{lemma}
	\label{lem:tree_correctness}	
	Let $ G $ be a simple graph and $ \Gamma $ a straight-line drawing of $ G $. If the intersection graph $ C = (V,E) $ of $ G $ is a tree with root $ r \in V $ and maximum degree $ k $, then $ \sollong(r) = \max \{T_i(r) \mid 0 \leq i \leq k \} $ is an optimal solution to the \maxsped problem on $ G $ with drawing $ \Gamma $.
\end{lemma}
\begin{proof}
	Let $ G $, $ \Gamma $ and $ C = (V,E) $ as above and $ u \in V $ a leaf of $ C $. Since $ c(u) $ is empty we simply get $ \solshort(u) = T_0(u) = \ell_1(u) $ and $ \sollong(u) = T_1(u) = \max\{\ell_0(u), \ell_1(u)\} $. Now let $ u \in V $ with degree $ \delta $ be an inner vertex. The values $ \solshort(v) $ and $ \sollong(v) $ are computed for all $ v \in c(u) $. $ T_i(u) $ for one $ 0 \leq i \leq \delta $ is then the sum of the maximum stubs we can draw among the children plus the stub length $ \ell_i(u) $ itself, which are just the values we are interested in. Setting then $ \sollong(u) $ and $ \solshort(u) $ as above gives the two necessary values for $ p(u) $.

%	and $ t = \sollong(r) $ is the maximum value for some $ T_i(r) $'s and $ r $ the root. Now let $ t' $ be the value of a solution to the \maxsped problem on $ G $ with drawing $ \Gamma $ and $ t' > t $. This solution has to be the sum of a set of stub lengths for each edge in $ G $ such that no two stubs intersect by definition. Further we know the stubs can be categorized into two sets relative to their parent. Recurrence~\ref{rec:tree} builds just the maximum over the $ \solshort(v) $ and $ \sollong(v) $ values for all $ v \in c(u) $, such that no two stubs intersect. Hence if $ t $ was not optimal we would need to find at least a third stub length for at least one child $ v \in c(u) $ which was neither equal to $ \sollong(v) $ nor $ \solshort(v) $.
\end{proof}

\begin{figure}
	\centering
	\includegraphics{speedup}
	\caption{The situation in Lemma~\ref{lem:compute_vertex}.}
	\label{fig:speedup}
\end{figure}
Solving Recurrence~\ref{rec:tree} straight forward would yield an algorithm running in $ O(mk^2) $. Using the order on the stub lengths though we can improve this to $ O(mk) $ by solving $ T_i(u) $ in linear time for one $ u \in V $.

\begin{lemma}
	\label{lem:compute_vertex}
	Let $ C = (V,E) $ be a tree with maximum degree $ k $ and vertex weights $ \ell_0,\dots,\ell_{\delta} $ for every $ u \in V $ with degree $ \delta $. We can compute $ \solshort(u) $ and $ \sollong(u) $ in $ O(\delta) $ time and space, if the values $ \sollong(v) $ and $ \solshort(v) $ are computed for all children $ v \in c(u) $.
\end{lemma}
\begin{proof}
	 Let $ u \in V $ be a vertex with degree $ \delta < k $ for which we have computed $ \sollong $ and $ \solshort $ for every of its children. The crucial part is that $ T_i(u) $ can be built iteratively by going through the list of stub lengths for $ u $ in ascending order. First we compute $ T_1(u) $, this is the solution for $ u $ in which we are able to draw the children with any of their stub lengths. This means we need to build the sum of $ \ell_1(u) $ and the $ \sollong(v) $ for every $ v \in c(u) $ which runs in $ O(\delta) $. The same can be done for $ T_0(u) $, but taking all the $ \solshort(v) $ values for the children $ v \in c(u) $. Assume we computed $ T_j(u) $ for all stub lengths $ \ell_i(u) $ with $ i \leq j $. Now we want to compute $ T_{j+1}(u) $ (see Figure~\ref{fig:speedup}). By drawing $ s(u) $ with stub length $ \ell_{j +1} $ we make it a bit longer since we consider the $ \ell_i(u) $ in ascending order. Hence by doing so we now potentially intersect the corresponding edge of the $ j $-th child $ v_j \in c(u) $. Since $ s(v_j) $ is the only segment affected by drawing $ u $ with stub length $ \ell_{j + 1} $ instead of $ \ell_j $ we can compute $ T_{j+1}(u) $ in $O(1)$ time by simply subtracting $ \sollong(v) $ and adding $ \solshort(v) $.

@@

\martin{the figures can still be improved in style and size; maybe merge into one, use nicer colors, draw just one segment and not multiple shifted ones etc}

@@

	Now let $ u \in V $ with degree $ \delta $ be an inner vertex of $ \mathcal T $. The values $ \solshort(v) $, $ \solmid(v) $ and $ \sollong(v) $ are computed for all $ v \in c(u) $. Again first let $ u $ be an articulation vertex, then the maximum for stub $ \ell_i(A(u)) $ is said value plus the sum of the compatible children, which is just Recurrence~\ref{rec:cactus_block}. Now let $ u $ be a block vertex. We have to cut open the cycle it represents (compare \cite{bcgkmn-pped-17}). We do this at $ p = A(p(u)) $. The resulting graph is a path over the vertices in $ \textit{cycle}(u) $ with attached articulation vertices. This is especially a tree and we use the aglorithm from Section~\ref{sec:tree}, maximizing over three values of the articulation vertices though,  to compute the $ T_i(u) $.
\end{proof}

@@

\begin{lemma}
	\label{lem:compute_block}
	Let $ \mathcal T = (\mathcal V,\mathcal E) $ be a cactus with maximum degree $ k $ and vertex weights $ \ell_0,\dots,\ell_{\delta} $ for every block vertex $ u \in \mathcal V $ with degree $ \delta $. We can compute $ \solshort(u) $, $ \solmid(u) $ and $ \sollong(u) $ in $ O(n\delta) $ time and space, if the values $ \sollong(v) $, $ \solmid(v) $ and $ \solshort(v) $ are computed for all children $ v \in c(u) $.
\end{lemma}
\begin{proof}
	Let $ G $, $ \Gamma $,  $ C = (V,E) $ and $ \mathcal T = (\mathcal V,\mathcal E) $ as above and $ u \in \mathcal V $ a block vertex with degree $ \delta $. Computing the value for the cycle itself $ \textit{cycle}(u) $ takes $ O(n) $ time by Bruckdorfer et al. Now at every vertex $ v \in \textit{cycle}(u) $ there might be a set of up to $ \delta $ many children, which are all articulation points. Their values are already known and maximization over them can hence be done in $ O(\delta) $ time by the same argument as in Lemma~\ref{lem:compute_vertex}. Following we need $ O(n\delta) $ to compute the three values $ \solshort(u) $, $ \solmid(u) $ and $ \sollong(u) $.
\end{proof}

\begin{theorem}
	Let $ G $ be a simple graph with $m$ edges and $ \Gamma $ a straight-line drawing of $ G $. If the intersection graph $ C = (V,E) $ of $ G $ is a cactus with maximum degree $ k \in \mathbb{N} $ the \maxsped problem can be solved in $ O(mk^2 + m \log m)) $ time and space.
\end{theorem}
\begin{proof}
	Follows as in Theorem~\ref{thm:tree} by maximizing over the tree using Lemma~\ref{lem:cactus_correctness} and Recurrence~\ref{rec:cactus_articulation}. Then the running time follows from constructing the intersection graph and the block-cut tree as well as Lemma~\ref{lem:compute_articulation} and~\ref{lem:compute_block}.
\end{proof}

>>>>

In this section we close the gap between the known hardness of \maxsped~\cite{b-sgh-15} and the polynomial-time algorithm for 2-plane drawings~\cite{bcgkmn-pped-17}.

@@

\begin{proof}
	We reduce from the \NP-hard problem \ppsat~\cite{l-pftu-82} using similar ideas as Bruckdorfer's general proof of the hardness of \maxsped~\cite{b-sgh-15}. 
	Let $\phi$ be a planar 3-Sat formula with $n$ variables $\{x_1, \dots, x_n\}$ and $m$ clauses $\{c_1, \dots, c_m\}$, each consisting of three literals.
	We can assume that $\phi$ comes with a planar drawing of its variable-clause graph $H_\phi$ that has a vertex for each variable $x_i$ placed on a horizontal line and a vertex for each clause $c_j$ connected to the three variables appearing in $c_j$. 
	In our gadget-based reduction we create a 3-plane drawing $\Gamma_\phi$ as a set of line segments of uniform length and a value $L$ such that $\Gamma$ has a SPED of total ink at least $L$ if and only if $\phi$ is satisfiable.

	All segments

@@

\end{proof}
\newpage %delete later just for editing purposes
\section{Two polynomial special cases}

@@

Section~\ref{sec:hardness} showed that \maxsped is generally \NP-hard for $k \ge 3$. In this section we consider the special case that the intersection graph of the $k$-plane input drawing is a tree or a cactus. In both cases we present polynomial-time dynamic programming algorithms. Let $ C = (V,E) $ be the intersection graph of a given drawing $\Gamma$ of graph $ G $ as defined in Section~\ref{sec:preliminaries}. Let $ u \in V $ and $ \deg(u) = \delta $, then for the corresponding edge $ s(u) \in S $ there are $ k + 1 $ possible stub pairs including drawing the whole edge. We store the lengths of the possible stubs as node weights $ \ell_1(u), \dots, \ell_\delta(u) \in \mathbb{R}_+$ and assume that $ \ell_1(u),\dots,\ell_\delta(u) $ are sorted from shorter to longer stubs. We define  $ \ell_0(u) $ as the length of the whole edge $ s(u) $.

@@

\subsection{Trees}
\label{sec:tree}
\begin{figure}[tbp]
	\centering
	\includegraphics{tree_illustration}
	\caption{One edge $ u $ with intersecting segments. The shaded areas are the stubs, $ p(u) $ is the parent of $ u $. The stubs in the boxed area contribute to $ \solshort(u) $.}
	\label{fig:tree}
\end{figure}
Here we assume that the intersection graph $ C = (V,E) $ is a rooted tree of maximum degree $ k $. We give a bottom up dynamic programming algorithm for solving \maxsped on $ C $. For a vertex $ u \in V $ let $ p(u) $ denote the parent of $ u $ in $ C $ and let $ c(u) $ denote the set of its children. For a vertex $ u \in V $ with parent $ p(u) \in V $ let $ a \in \mathbb{R}_+ $ be the shorter length between one of $ s(u) $'s endpoint and the intersection point between $ s(u) $ and $ s(p(u)) $. Now we can define the following two values, allowing us to categorize the stub-lengths into those not affecting the parents choice of stub and those which don't:
%Since we know $ C $ is a tree every vertex $ u \in V $ with degree $ \delta $ has a parent $ p(u) \in V $. For this parent there exists one stub length $ \ell_j(p(u)) $ in the sequence of stub lengths $ \ell_0(p(u)),\dots,\ell_{\delta}(p(u)) $ such that drawing $ p(u) $ with these stubs does not intersect a completely drawn $ u $, while drawing $ p(u) $ with $ \ell_{j + 1}(p(u)) $, i.e. the next longer stub length, intersects $ \ell_0(u) $. We call the solution which would intersect the parent $ \sollong(u) $ and the one not intersecting the parent $ \solshort(u) $. See Figure~\ref{fig:tree} for an illustration. More formally let $ j $ be the index such that stubs of length $ \ell_j(p(u)) $ do not intersect $ s(u) $, but $ \ell_{j +1}(p(u)) $ would, we set:
\begin{align*}
	\solshort(u) = \max\{T_1(u),\dots,T_a(u)\} &&
	\sollong(u) = \max\{T_0(u),\dots,T_{\delta}(u)\}
\end{align*}
It remains to define $ T_i(u) $ for a vertex $ u $ and stub length $ \ell_i(u) $:
\begin{align}
	\label{rec:tree}
	T_i(u) = \ell_i(u) + \sum_{v \in c(u)}
	\begin{cases}
		\solshort(v) & \text{if } s(u) \text{ with length } \ell_i \text{ intersects } s(v) \\
		\sollong(v) & \text{else}
	\end{cases}
\end{align}
Solving Recurrence~\ref{rec:tree} yields an optimal solution to the \maxsped problem on $ G $ with drawing $ \Gamma $ as the following lemma shows.

\begin{lemma}
	\label{lem:tree_correctness}	
	Let $ G $ be a simple graph and $ \Gamma $ a straight-line drawing of $ G $. If the intersection graph $ C = (V,E) $ of $ G $ is a tree with root $ r \in V $ and maximum degree $ k $, then $ \sollong(r) = \max \{T_i(r) \mid 0 \leq i \leq k \} $ is an optimal solution to the \maxsped problem on $ G $ with drawing $ \Gamma $.
\end{lemma}
\begin{proof}
	Let $ G $, $ \Gamma $ and $ C = (V,E) $ as above and $ u \in V $ a leaf of $ C $. Since $ c(u) $ is empty we simply get $ \solshort(u) = T_0(u) = \ell_1(u) $ and $ \sollong(u) = T_1(u) = \max\{\ell_0(u), \ell_1(u)\} $. Now let $ u \in V $ with degree $ \delta $ be an inner vertex. The values $ \solshort(v) $ and $ \sollong(v) $ are computed for all $ v \in c(u) $. $ T_i(u) $ for one $ 0 \leq i \leq \delta $ is then the sum of the maximum stubs we can draw among the children plus the stub length $ \ell_i(u) $ itself, which are just the values we are interested in. Setting then $ \sollong(u) $ and $ \solshort(u) $ as above gives the two necessary values for $ p(u) $.
%	and $ t = \sollong(r) $ is the maximum value for some $ T_i(r) $'s and $ r $ the root. Now let $ t' $ be the value of a solution to the \maxsped problem on $ G $ with drawing $ \Gamma $ and $ t' > t $. This solution has to be the sum of a set of stub lengths for each edge in $ G $ such that no two stubs intersect by definition. Further we know the stubs can be categorized into two sets relative to their parent. Recurrence~\ref{rec:tree} builds just the maximum over the $ \solshort(v) $ and $ \sollong(v) $ values for all $ v \in c(u) $, such that no two stubs intersect. Hence if $ t $ was not optimal we would need to find at least a third stub length for at least one child $ v \in c(u) $ which was neither equal to $ \sollong(v) $ nor $ \solshort(v) $.
\end{proof}

Solving Recurrence~\ref{rec:tree} straight forward would yield an algorithm running in $ O(mk^2) $. Using the order on the stub lengths though we can improve this to $ O(mk) $ by solving $ T_i(u) $ in linear time for one $ u \in V $.

\begin{lemma}
	\label{lem:compute_vertex}
	Let $ C = (V,E) $ be a tree with maximum degree $ k $ and vertex weights $ \ell_0,\dots,\ell_{\delta} $ for every $ u \in V $ with degree $ \delta $. We can compute $ \solshort(u) $ and $ \sollong(u) $ in $ O(\delta) $ time and space, if the values $ \sollong(v) $ and $ \solshort(v) $ are computed for all children $ v \in c(u) $.
\end{lemma}
\begin{proof}
	 Let $ u \in V $ be a vertex with degree $ \delta < k $ for which we have computed $ \sollong $ and $ \solshort $ for every of its children. The crucial part is that $ T_i(u) $ can be built iteratively by going through the list of stub lengths for $ u $ in ascending order. First we compute $ T_1(u) $, this is the solution for $ u $ in which we are able to draw the children with any of their stub lengths. This means we need to build the sum of $ \ell_1(u) $ and the $ \sollong(v) $ for every $ v \in c(u) $ which runs in $ O(\delta) $. The same can be done for $ T_0(u) $, but taking all the $ \solshort(v) $ values for the children $ v \in c(u) $. Assume we computed $ T_j(u) $ for all stub lengths $ \ell_i(u) $ with $ i \leq j $. Now we want to compute $ T_{j+1}(u) $ (see Figure~\ref{fig:tree}). By drawing $ s(u) $ with stub length $ \ell_{j +1} $ we make it a bit longer since we consider the $ \ell_i(u) $ in ascending order. Hence by doing so we now potentially intersect the corresponding edge of the $ j $-th child $ v_j \in c(u) $. Since $ s(v_j) $ is the only segment affected by drawing $ u $ with stub length $ \ell_{j + 1} $ instead of $ \ell_j $ we can compute $ T_{j+1}(u) $ in $O(1)$ time by simply subtracting $ \sollong(v) $ and adding $ \solshort(v) $.

@@

@@

	Now let $ u \in V $ with degree $ \delta $ be an inner vertex of $ \mathcal T $. The values $ \solshort(v) $, $ \solmid(v) $ and $ \sollong(v) $ are computed for all $ v \in c(u) $. Again first let $ u $ be an articulation vertex, then the maximum for stub $ \ell_i(A(u)) $ is said value plus the sum of the compatible children, which is just Recurrence~\ref{rec:cactus_articulation}. Now let $ u $ be a block vertex. We have to cut open the cycle it represents (compare \cite{bcgkmn-pped-17}). We do this at $ p = A(p(u)) $. The resulting graph is a path over the vertices in $ \textit{cycle}(u) $ with attached articulation vertices. This is especially a tree and we use the aglorithm from Section~\ref{sec:tree}, maximizing over three values of the articulation vertices though,  to compute the $ T_i(u) $.
\end{proof}

@@

\begin{lemma}
	\label{lem:compute_block}
	Let $ \mathcal T = (\mathcal V,\mathcal E) $ be a cactus with maximum degree $ k $ and vertex weights $ \ell_0,\dots,\ell_{\delta} $ for every block vertex $ u \in \mathcal V $ with degree $ \delta $. We can compute $ \solshort(u) $, $ \solmid(u) $ and $ \sollong(u) $ in $ O(n\delta) $ time and space, if the values $ \sollong(v) $, $ \solmid(v) $ and $ \solshort(v) $ are computed for all children $ v \in c(u) $.
\end{lemma}
\begin{proof}
	Let $ G $, $ \Gamma $,  $ C = (V,E) $ with maximum degree $ k $ and $ \mathcal T = (\mathcal V,\mathcal E) $ as above and $ u \in \mathcal V $ a block vertex. Computing the value for the cycle itself $ \textit{cycle}(u) $ takes $ O(n_c) $ time by Bruckdorfer et al. with $ n_c $ being the number of vertices in the cycle. Now at every vertex $ v \in \textit{cycle}(u) $ there might be a set of up to $ k $ many children, which are all articulation points. Their values are already known and maximization over them can hence be done in $ O(k) $ time by the same argument as in Lemma~\ref{lem:compute_vertex}. Following we need $ O(n_ck) $ to compute the three values $ \solshort(u) $, $ \solmid(u) $ and $ \sollong(u) $.
\end{proof}

\begin{theorem}
	Let $ G $ be a simple graph with $m$ edges and $ \Gamma $ a straight-line drawing of $ G $. If the intersection graph $ C = (V,E) $ of $ G $ is a cactus with maximum degree $ k \in \mathbb{N} $ the \maxsped problem can be solved in $ O(mk + m \log m)) $ time and space.
\end{theorem}
\begin{proof}
	Follows as in Theorem~\ref{thm:tree} by maximizing over the tree using Lemma~\ref{lem:cactus_correctness} and Recurrence~\ref{rec:cactus_articulation}. Then the running time follows from constructing the intersection graph and the block-cut tree as well as Lemma~\ref{lem:compute_articulation} and~\ref{lem:compute_block}.
\end{proof}
